---
title: "third_pass_elvin"
author: "Elvin Liu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup}
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(labelled)
library(Matrix)
library(glmnet)
```

```{r}
# map labels from txt to csv
lines <- readLines("J362221_labels.txt")
var_lines <- lines[grepl("^ER[0-9A-Z]+\\s+", lines)]

label_df <- data.frame(
  var   = trimws(sub("^(ER[0-9A-Z]+)\\s+.*$", "\\1", var_lines)),
  label = trimws(sub("^ER[0-9A-Z]+\\s+(.*?)\\s{2,}\\d{0,2}\\s*$", "\\1", var_lines)),
  stringsAsFactors = FALSE
)

df <- read.csv("J362221.csv", stringsAsFactors = FALSE)

for (i in seq_len(nrow(label_df))) {
  v <- label_df$var[i]
  if (v %in% names(df)) {
    var_label(df[[v]]) <- label_df$label[i]
  }
}

# get rid of dupe family heads
df <- df %>%
  group_by(ER10002) %>%
  slice(1) %>%
  ungroup()

# convert NA values for 0, 8, and 9 values for TANF
df$ER11272[df$ER11272 == 0 | df$ER11272 == 8 | df$ER11272 == 9] <- NA

# remove NA values for TANF
df <- df[!is.na(df$ER11272), ]

# do the same manually for VA
# 1-4 are treated as 1, 8 and 9 as VA
# VA whether received variables
va_vars <- c(
  "ER11319", "ER14585", "ER18747", "ER22119",
  "ER26100", "ER37118", "ER43109"
)

df[va_vars] <- lapply(df[va_vars], function(x) {
  x[x %in% c(1,2,3,4)] <- 1
  x[x %in% c(8,9)] <- NA
  x
})

# 1 is Yes, 5 is No, TANF
df <- df[df$ER11272 == 1 | df$ER11272 == 5,]

# send 999999, 999998, 99999, 99998 to NA
df[df == 999999 | df == 999998 | df == 99999 | df == 99998] <- NA

# ER41027 <- Total Family Income 2007
# ER41029 <- Census Needs Standard 2007
# ER12079 <- TFI 1997
# ER12220 <- CNS 1997

# calculate 1997 TFI / CNS ratio
df$income_to_needs_ratio_1997 <- df$ER12079 / df$ER12220

# calculate 2007 TFI / CNS ratio
df$income_to_needs_ratio_2007 <- df$ER41027 / df$ER41029

# keep only people with 1997 and 2007 ratio
df <- df %>%
  filter(!is.na(income_to_needs_ratio_1997) & !is.na(income_to_needs_ratio_2007))

# kick out TANF non-recipients with too high of income or no kids or too old
income_ratio_threshold <- quantile(df[df$ER11272 == 1,]$income_to_needs_ratio_1997, 0.90) # 95th percentile income of TANF recipients
df <- df[(df$ER11272 == 5 & df$income_to_needs_ratio_1997 <= income_ratio_threshold & df$ER10012 >= 1 & df$ER10009 < 65) | df$ER11272 == 1,]

# 233 families who received TANF in 1997 and have both 1997 and 2007 ratio
# 609 families who didn't receive TANF in 1997 and have both 1997 and 2007 ratio

# mapping from wide to long
wave_map <- list(
  "1997" = list(
    # Tracking IDs
    family_id        = "ER10002",
    person_id        = "ER33401",
    
    income_to_needs_ratio = "income_to_needs_ratio_1997",
    
    # Demographics & Geography
    state            = "ER12221",
    age              = "ER10009",
    sex              = "ER10010",
    n_children       = "ER10012",
    age_yng_child    = "ER10013",
    race             = "ER11848",
    education        = "ER12222",
    couple_status    = "ER12223C",
    relation         = "ER33403",
    emp_status       = "ER33411",
    rural_urban      = "ER12221A",
    
    # Housing & Environment
    type_du          = "ER10017",
    own_rent         = "ER10035",
    neither_own_rent = "ER10068",
    food_stamps      = "ER11065",
    
    # Health
    health_head      = "ER11723",
    health_wife      = "ER11727",
    parents_poor     = "ER11846",
    
    # Income & Needs
    total_income     = "ER12079",
    needs_standard   = "ER12220",
    
    # Transfer Income
    wtr_adc          = "ER11272",
    amt_adc          = "ER11273",
    wtr_ssi          = "ER11287",
    amt_ssi          = "ER11289",
    wtr_welfare      = "ER11303",
    amt_welfare      = "ER11304",
    wtr_va           = "ER11319",
    amt_va           = "ER11322",
    wtr_nonva_ret    = "ER11336",
    amt_nonva_ret    = "ER11337",
    wtr_annuities    = "ER11351",
    amt_annuities    = "ER11352",
    wtr_otr_pen      = "ER11366",
    amt_otr_pen      = "ER11367",
    wtr_unemp        = "ER11382",
    amt_unemp        = "ER11383",
    wtr_workers_comp = "ER11397",
    amt_workers_comp = "ER11398",
    wtr_child_supp   = "ER11412",
    amt_child_supp   = "ER11413"
  ),
  "2007" = list(
    # Tracking IDs
    family_id        = "ER36002",
    person_id        = "ER33901",
    
    income_to_needs_ratio = "income_to_needs_ratio_2007",
    
    # Demographics & Geography
    state            = "ER36003",
    age              = "ER36017",
    sex              = "ER36018",
    n_children       = "ER36020",
    age_yng_child    = "ER36021",
    race             = "ER40565",
    education        = "ER41037",
    couple_status    = "ER41041",
    relation         = "ER33903",
    emp_status       = "ER33913",
    rural_urban      = "ER41033",
    
    # Housing & Environment
    type_du          = "ER36024",
    own_rent         = "ER36028",
    neither_own_rent = "ER36075",
    food_stamps      = "ER36701",
    
    # Health
    health_head      = "ER38202",
    health_wife      = "ER39299",
    parents_poor     = "ER37728",
    
    # Income & Needs
    total_income     = "ER41027",
    needs_standard   = "ER41029",
    
    # Transfer Income
    wtr_adc          = "ER37068",
    amt_adc          = "ER37069",
    wtr_ssi          = "ER37084",
    amt_ssi          = "ER37086",
    wtr_welfare      = "ER37101",
    amt_welfare      = "ER37102",
    wtr_va           = "ER37118",
    amt_va           = "ER37119",
    wtr_nonva_ret    = "ER37134",
    amt_nonva_ret    = "ER37135",
    wtr_annuities    = "ER37150",
    amt_annuities    = "ER37151",
    wtr_otr_pen      = "ER37166",
    amt_otr_pen      = "ER37167",
    wtr_unemp        = "ER37183",
    amt_unemp        = "ER37184",
    wtr_workers_comp = "ER37199",
    amt_workers_comp = "ER37200",
    wtr_child_supp   = "ER37215",
    amt_child_supp   = "ER37216"
  )
)

# map from wide to long
df_long <- names(wave_map) %>%
  map_dfr(function(year_str) {
    
    year_mapping <- wave_map[[year_str]]
    
    # Remove person_id and family_id from the loop mapping to prevent tracking splits
    year_mapping$person_id <- NULL
    year_mapping$family_id <- NULL
    
    rename_vec <- unlist(year_mapping)
    rename_vec <- rename_vec[rename_vec %in% names(df)]
    select_vec <- setNames(rename_vec, names(rename_vec))
    
    # Extract the correct interview number code before stepping into mutate
    current_wave_num <- if (year_str == "1997") df$ER10002 else df$ER36002
    
    df %>%
      select(all_of(select_vec)) %>%
      mutate(
        year = as.integer(year_str),
        # Anchor both waves to the unique 1997 Family ID row number
        unique_family_id = df$ER10002,
        # Assign the pre-extracted vector safely
        wave_interview_num = current_wave_num
      )
  })

# Create a clean, time-invariant mapping of 1997 TANF status using your original df
tanf_status_map <- df %>%
  mutate(tanf_status = if_else(ER11272 == 1, "TANF Recipient", "Control Group")) %>%
  select(unique_family_id = ER10002, tanf_status)

# Merge this structural indicator back into your long-form panel
df_long <- df_long %>%
  left_join(tanf_status_map, by = "unique_family_id")

# Move your unified tracking key and year metrics to the front
df_long <- df_long %>%
  select(unique_family_id, wave_interview_num, year, tanf_status, income_to_needs_ratio, everything())
```

```{r}
# Verify every panel tracker contains exactly 2 observations
panel_counts <- df_long %>% 
  count(unique_family_id)

table(panel_counts$n) 
# This should show 100% of your records sit in the "2" category!
```

```{r}
#Initial Visualization: 2 line plot of group means over time

plot_data <- df_long %>%
  group_by(year, tanf_status) %>%
  summarize(
    mean_ratio = mean(income_to_needs_ratio, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(plot_data,
       aes(x = year,
           y = mean_ratio,
           color = tanf_status,
           group = tanf_status)) +
  geom_line(linewidth = 1.2) +
  geom_point(linewidth = 3) +
  labs(
    title = "Income-to-Needs Ratio by TANF Status",
    x = "Year",
    y = "Income-to-Needs Ratio",
    color = ""
  ) +
  theme_minimal()

```

```{r}
#Model 1: Difference in Difference Regression for change in Income-Needs ratio for Tanf and Non-Tanf Recipients (No Controls)

#Create Post indicator to verify which year the family being observed is in 
df_long$post <- ifelse(df_long$year == 2007, 1, 0)

#Ensure that TANF status is a factor
df_long$tanf_status <- factor(df_long$tanf_status)
levels(df_long$tanf_status)

#Run the DiD Regression: 
model1 <- lm(
  income_to_needs_ratio ~
    tanf_status * post,
  data = df_long
)

summary(model1)


```
$$
IncomeNeed_{it}
=
\beta_0
+
\beta_1 TANF_i
+
\beta_2 Post_t
+
\beta_3 (TANF_i \times Post_t)
+
\varepsilon_{it}
$$ 
The primary coefficient of interest is $\beta_3$, which measures whether TANF recipients experienced a different change in income-to-needs ratios between 1997 and 2007 relative to the control group. From this result, we can see that TANF recipients started siginificantly poorer than the control group, with a Income to Needs Ratio .26123 Less than the control group. The control group improved their ratio by 1.25172 over the time period measured whereas TANF recipients improved their Income to Needs Ratio by .55706 points less than those in the control groups. All results are statistically significant. 

```{r}
#Model 2: Run a DiD regression controlling for differences in age, education, race, family size, types of welfare recieved, etc. 

#Data Cleaning On Control Variables. Get rid of 99 values for education, get red of 7, 8, 9 values for race, as these are NA, and for ease of LASSO in next step convert welfare variables to Binary 1=Yes and 0=No, make sure 8, 9, 0 values are changed to NA for these welfare variables. 

# Education: 99 = DK/NA
df_long$education[df_long$education == 99] <- NA

# Race: 8 and 9 = DK/NA
# Keep 1-7 as valid race categories
df_long$race <- as.character(df_long$race)

df_long$race[df_long$race %in% c("0", "8", "9")] <- NA

df_long$race <- factor(df_long$race)

# Safer yes/no cleaner
# Raw codebook: 1 = Yes, 5 = No, 0/8/9 = missing
# If already cleaned: 1 = Yes, 0 = No
clean_yesno_safe <- function(x) {
  x <- as.character(x)

  x[x == "1"] <- "1"
  x[x == "5"] <- "0"
  x[x %in% c("8", "9")] <- NA

  as.numeric(x)
}

binary_vars <- c(
  "wtr_ssi",
  "wtr_welfare",
  "wtr_unemp",
  "wtr_child_supp",
  "wtr_workers_comp",
  "wtr_va"
)

df_long[binary_vars] <- lapply(df_long[binary_vars], clean_yesno_safe)

# Check cleaning
summary(df_long[, c(
  "age",
  "education",
  "race",
  "n_children",
  binary_vars
)])

sapply(df_long[binary_vars], table, useNA = "always")
table(df_long$race, useNA = "always")
```
```{r}
#Run Model 2 with the cleaned data and binary values
model2 <- lm(
  income_to_needs_ratio ~
    tanf_status * post +
    age +
    education +
    race +
    n_children +
    wtr_ssi +
    wtr_welfare +
    wtr_unemp +
    wtr_child_supp +
    wtr_workers_comp +
    wtr_va,
  data = df_long
)

summary(model2)
nobs(model2)
```
$$
IncomeNeed_{it}
=
\beta_0
+
\beta_1 TANF_i
+
\beta_2 Post_t
+
\beta_3 (TANF_i \times Post_t)
+
\beta_4 Age_i
+
\beta_5 Education_i
+
\beta_6 Children_i
+
\beta_7 SSI_i
+
\beta_8 Welfare_i
+
\beta_9 Unemp_i
+
\beta_{10} ChildSupp_i
+
\beta_{11} WorkersComp_i
+
\beta_{12} VA_i
+
\sum_r \gamma_r Race_{ir}
+
\varepsilon_{it}
$$
After controlling for age, education, race, number of children, SSI, welfare income, unemployment compensation, child support, workers compensation, and VA benefits, TANF recipients experienced an increase in income-to-needs ratio that was approximately 0.56 smaller than the increase experienced by comparable nonrecipients between 1997 and 2007. The $\beta_3$ coefficient only changed by about .00078 which is a negligible difference. This indicates that our selection of the control group is very robust, and that the negative TANF effect is being significantly effected by the control variables selected. From looking at the P-values of the control variables, education, number of children, SSI, and Race2 (African American Individuals). 

```{r}
#Model 3: Now controlling for 16 variables, use a double LASSO to verify the Robustness of the Model. This will use the data to predict which of the 16 possible co-founding variables have the largest effect on the change of the income to needs ratio between TANF recipients and NON-TANF recipients. 

# Model 3: Double LASSO robustness check

# Check variables for data cleaning
vars_to_check <- c(
  "age",
  "education",
  "race",
  "n_children",
  "age_yng_child",
  "couple_status",
  "emp_status",
  "food_stamps",
  "health_head",
  "parents_poor",
  "wtr_ssi",
  "wtr_welfare",
  "wtr_unemp",
  "wtr_child_supp",
  "wtr_workers_comp",
  "wtr_va"
)

lapply(df_long[vars_to_check], table, useNA = "always")


# Additional cleaning for Model 3 only

# Couple status: categorical
df_long$couple_status <- factor(df_long$couple_status)

# Employment status: 8 = Other, 9 = DK/NA
df_long$emp_status[df_long$emp_status %in% c(8, 9)] <- NA
df_long$emp_status <- factor(df_long$emp_status)

# Food stamps: 0 = no food stamps, 1-7 = number of family members covered, 8/9 = DK/NA
df_long$food_stamps[df_long$food_stamps %in% c(8, 9)] <- NA

# Health status of head: 1 = Excellent ... 5 = Poor, 8/9 = DK/NA
df_long$health_head[df_long$health_head %in% c(8, 9)] <- NA

# Parents poor: 1 = Poor, 3 = Average, 5 = Well-off, 0/8/9 = invalid or DK/NA
df_long$parents_poor[df_long$parents_poor %in% c(0, 8, 9)] <- NA

# Check cleaned Model 3 variables
lapply(df_long[vars_to_check], table, useNA = "always")
```
```{r}
#Model 3: Now do the double LASSO: 

#Create a vector of all the con founders
candidate_controls <- c(
  "age",
  "education",
  "race",
  "n_children",
  "age_yng_child",
  "couple_status",
  "emp_status",
  "food_stamps",
  "health_head",
  "parents_poor",
  "wtr_ssi",
  "wtr_welfare",
  "wtr_unemp",
  "wtr_child_supp",
  "wtr_workers_comp",
  "wtr_va"
)

#Give Numeric Treatment indicator for Double Lasso
df_lasso <- df_long

df_lasso$treat <- ifelse(
  df_lasso$tanf_status == "TANF Recipient",
  1,
  0
)

#Remove missing variables from data
lasso_data <- df_lasso[, c(
  "income_to_needs_ratio",
  "treat",
  "post",
  "age",
  "education",
  "race",
  "n_children",
  "age_yng_child",
  "couple_status",
  "emp_status",
  "food_stamps",
  "health_head",
  "parents_poor",
  "wtr_ssi",
  "wtr_welfare",
  "wtr_unemp",
  "wtr_child_supp",
  "wtr_workers_comp",
  "wtr_va"
)]

lasso_data <- na.omit(lasso_data)

#Create Variables for LASSO
Y <- lasso_data$income_to_needs_ratio

D <- lasso_data$treat

#Factor variables converted to Dummy Variables for Ease of the LASSO
X <- model.matrix(
  ~ age +
    education +
    race +
    n_children +
    age_yng_child +
    couple_status +
    emp_status +
    food_stamps +
    health_head +
    parents_poor +
    wtr_ssi +
    wtr_welfare +
    wtr_unemp +
    wtr_child_supp +
    wtr_workers_comp +
    wtr_va,
  data = lasso_data
)[,-1]
```

```{r}
set.seed(424)
#First Lasso: which variables are important for predicting change in income to needs ratio
cv_y <- cv.glmnet(
  X,
  Y,
  alpha = 1
)

#Second Lasso: which variables are important in predicting who recieves TANF
cv_d <- cv.glmnet(
  X,
  D,
  alpha = 1,
  family = "binomial"
)

#Extract Selected variables in the First LASSO
selected_y <- rownames(coef(cv_y, s = "lambda.min"))[
  coef(cv_y, s = "lambda.min")[,1] != 0
]

#Extract Selected variables in the Second LASSO
selected_d <- rownames(coef(cv_d, s = "lambda.min"))[
  coef(cv_d, s = "lambda.min")[,1] != 0
]

# Create a Union of the Selected Variables
selected_controls <- union(selected_y, selected_d)

selected_controls <- setdiff(
  selected_controls,
  "(Intercept)"
)

# Show selected dummy-variable columns
selected_controls

# Convert dummy variables back to original variables
selected_original <- unique(
  gsub("[0-9]+$", "", selected_controls)
)

selected_original

# Number of selected columns
length(selected_controls)

# Percent of columns kept
percent_kept <- length(selected_controls) / ncol(X)
percent_kept

```
```{r}
#Final Step: create the final regression using only LASSO selected controls
# Create formula using selected controls

formula_lasso <- as.formula(
  paste(
    "income_to_needs_ratio ~ treat * post +",
    paste(selected_original, collapse = " + ")
  )
)

formula_lasso
```

```{r}
#Run the Regresion
model3 <- lm(
  formula_lasso,
  data = lasso_data
)

summary(model3)
```
"TANF recipients experienced a 0.89 lower change in income-to-needs ratios between 1997 and 2007 relative to comparable nonrecipients, conditional on the selected controls."

```{r}
#Run Model 2 on the LASSO sample to check if the different in treat:post is due to lost observations
model2_same_sample <- lm(
  income_to_needs_ratio ~
    treat * post +
    age +
    education +
    race +
    n_children +
    wtr_ssi +
    wtr_welfare +
    wtr_unemp +
    wtr_child_supp +
    wtr_workers_comp +
    wtr_va,
  data = lasso_data
)

summary(model2_same_sample)
```
From these regressions, we notice that in the DOUBLE LASSO model, the coeeficient of interest $\beta_3 changes from -.557 to -.89. We run a test on the same observations from the LASSO by using model 2, and the coeeficient is -.58. This indicates that the jump in the coeeficient to -.89 in the double LASSO model can largely be explained by added variables that the LASSO includes such as employment status, marital status, health, food stamp usage, and child economic background. 

Across all specifications, TANF recipients experienced significantly smaller improvements in income-to-needs ratios between 1997 and 2007 than comparable nonrecipients. The estimated effect ranged from approximately -0.56 to -0.89 depending on the specification. The consistency of the negative treatment effect across baseline, controlled, and Double LASSO models suggests that the result is not driven by a particular choice of controls.


```{r}
# graphs of data!
```